---
title: "End-to-end coralseed workflow: Lizard Island"
author: "George Roff"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This tutorial runs the core `coralseed` workflow for a simulated reseeding event at Lizard Island:

1. load reef and benthic habitat maps;
2. convert habitat classes into a spatial seascape settlement-probability surface;
3. seed particles from an external dispersal model;
4. simulate settlement;
5. estimate settlement density and settlement summaries;
6. visualise the output.


```{r}
library(coralseed)
library(tidyverse)
library(sf)
library(tmap)
library(ggplot2)

sf::sf_use_s2(FALSE)
```

## Load example spatial data

The package includes Lizard Island example data in `inst/extdata`. These are accessed with `system.file()` so the tutorial works after installation.

```{r}
lizard_benthic_map <- system.file(
  "extdata", "Lizard_Benthic.geojson",
  package = "coralseed"
) |>
  sf::st_read(quiet = TRUE)

lizard_reef_map <- system.file(
  "extdata", "Lizard_Geomorphic.geojson",
  package = "coralseed"
) |>
  sf::st_read(quiet = TRUE)

lizard_particles_sf <- system.file(
  "extdata", "lizard_del_14_1512_sim1_10.json",
  package = "coralseed"
) |>
  sf::st_read(quiet = TRUE)
```

## Build the seascape probability surface

`seascape_probability()` combines reef geomorphology and benthic habitat maps and assigns settlement probabilities to habitat classes.

```{r}
lizard_seascape <- seascape_probability(
  reefoutline = lizard_reef_map,
  habitat = lizard_benthic_map
)

lizard_seascape
```

```{re}
ggplot(lizard_seascape) +
  geom_sf(aes(fill = settlement_probability), colour = NA) +
  scale_fill_viridis_c(na.value = "grey90") +
  labs(
    fill = "Settlement\nprobability",
    title = "Lizard Island settlement-probability seascape"
  ) +
  theme_bw()
```

## Seed particles

`seed_particles()` takes particle positions from an external dispersal model and adds biological filters, including time limits, competency, mortality, and settlement probability.

```{r}
lizard_particles <- seed_particles(
  input = lizard_particles_sf,
  zarr = FALSE,
  set.centre = TRUE,
  seascape = lizard_seascape,
  probability = "additive",
  limit.time = 12,
  competency.function = "exponential",
  crs = 20353,
  simulate.mortality = "typeIII",
  simulate.mortality.n = 0.1,
  return.plot = TRUE,
  return.summary = TRUE,
  silent = FALSE
)
```

Inspect the returned object.

```{r}
names(lizard_particles)

lizard_particles$summary
```

## Settle particles

`settle_particles()` applies spatial settlement probability and returns the simulated settled subset.

```{r}
lizard_settlers <- settle_particles(
  lizard_particles,
  probability = "additive",
  return.plot = FALSE,
  silent = TRUE
)

names(lizard_settlers)
```

## Plot settled particles

```{r}
#| fig-height: 7
#| fig-width: 9
plot_particles(
  lizard_settlers$points,
  lizard_seascape
)
```

## Estimate settlement density

`settlement_density()` converts settled particles into a spatial density surface.

```{r}
lizard_settlement_density <- settlement_density(
  lizard_settlers$points
)

lizard_settlement_density
```

## Summarise settlement

`settlement_summary()` gives spatially explicit summary metrics, usually binned to a selected cell size.

```{r}
lizard_settlement_summary <- settlement_summary(
  lizard_particles,
  lizard_settlers,
  cellsize = 50
)

lizard_settlement_summary
```

## Interactive map

Interactive leaflet widgets can be heavy in pkgdown builds. For stable documentation, the live `map_coralseed()` call is shown but not evaluated by default.

```{r}
#| eval: false
map_coralseed(
  seed_particles_input = lizard_particles,
  settle_particles_input = lizard_settlers,
  settlement_density_input = lizard_settlement_density,
  seascape_probability = lizard_seascape,
  restoration.plot = c(100, 100),
  show.footprint = TRUE,
  show.tracks = TRUE,
  subsample = 1000,
  webGL = TRUE
)
```

To save a stable HTML widget for pkgdown:

```{r}
#| eval: false
lizard_map <- map_coralseed(
  seed_particles_input = lizard_particles,
  settle_particles_input = lizard_settlers,
  settlement_density_input = lizard_settlement_density,
  seascape_probability = lizard_seascape,
  restoration.plot = c(100, 100),
  show.footprint = TRUE,
  show.tracks = TRUE,
  subsample = 1000,
  webGL = TRUE
)

fs::dir_create("vignettes/www")
htmlwidgets::saveWidget(
  lizard_map,
  "vignettes/www/lizard_map.html",
  selfcontained = TRUE
)
```

Then embed it manually:

```{=html}
<!-- <iframe src="www/lizard_map.html" width="100%" height="700"></iframe> -->
```

## Model flowchart

```{r flowchart}
#| fig-height: 8
#| fig-width: 10
flowchart_coralseed(
  lizard_particles,
  lizard_settlers,
  multiplier = 1000,
  postsettlement = 0.8
)
```

## Key outputs

```{r}
list(
  seascape = lizard_seascape,
  seeded_particles = lizard_particles,
  settlers = lizard_settlers,
  settlement_density = lizard_settlement_density,
  settlement_summary = lizard_settlement_summary
)
```
