---
title: "Build settlement-probability seascapes"
author: "George Roff"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This tutorial focuses on the spatial habitat component of `coralseed`: converting reef and benthic maps into a settlement-probability seascape.

```{r}
library(coralseed)
library(tidyverse)
library(sf)
library(ggplot2)

sf::sf_use_s2(FALSE)
```

## Load example maps

```{r}
benthic_map <- system.file("extdata", "Lizard_Benthic.geojson", package = "coralseed") |>
  st_read(quiet = TRUE)

reef_map <- system.file("extdata", "Lizard_Geomorphic.geojson", package = "coralseed") |>
  st_read(quiet = TRUE)
```

## Inspect habitat classes

```{r}
benthic_map |>
  st_drop_geometry() |>
  glimpse()

reef_map |>
  st_drop_geometry() |>
  glimpse()
```

```{r}
benthic_map |>
  st_drop_geometry() |>
  count(class, sort = TRUE)
```

Adjust the `class` column name if the source dataset uses a different habitat field.

## Create a seascape probability layer

```{r}
seascape <- seascape_probability(
  reefoutline = reef_map,
  habitat = benthic_map
)

seascape
```

## Visualise the seascape

```{r}
#| fig-height: 7
#| fig-width: 9

ggplot(seascape) +
  geom_sf(aes(fill = settlement_probability), colour = NA) +
  scale_fill_viridis_c(na.value = "grey85") +
  labs(
    title = "Habitat-specific settlement probability",
    fill = "Settlement\nprobability"
  ) +
  theme_bw()
```

## Summarise available settlement habitat

```{r}
seascape_area <- seascape |>
  st_transform(20353) |>
  mutate(area_m2 = as.numeric(st_area(geometry))) |>
  st_drop_geometry() |>
  group_by(settlement_probability) |>
  summarise(
    area_m2 = sum(area_m2, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    area_ha = area_m2 / 10000,
    prop_area = area_m2 / sum(area_m2, na.rm = TRUE)
  )

seascape_area
```

```{r}
ggplot(seascape_area, aes(factor(settlement_probability), area_ha)) +
  geom_col() +
  labs(
    x = "Settlement probability",
    y = "Area (ha)",
    title = "Available area by settlement-probability class"
  ) +
  theme_bw()
```

## Export the seascape

```{r}
#| eval: false
st_write(
  seascape,
  "outputs/lizard_seascape_probability.gpkg",
  delete_dsn = TRUE
)
```
