---
title: "Seed particles, competency, and mortality"
author: "George Roff"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This tutorial focuses on `seed_particles()`: the step that converts particle-track output into biologically filtered larval particles by applying dispersal-time limits, competency, mortality, and habitat settlement probability.

```{r}
library(coralseed)
library(tidyverse)
library(sf)
library(ggplot2)

sf::sf_use_s2(FALSE)
```

## Prepare seascape and particle inputs

```{r}
benthic_map <- system.file("extdata", "Lizard_Benthic.geojson", package = "coralseed") |>
  st_read(quiet = TRUE)

reef_map <- system.file("extdata", "Lizard_Geomorphic.geojson", package = "coralseed") |>
  st_read(quiet = TRUE)

seascape <- seascape_probability(
  reefoutline = reef_map,
  habitat = benthic_map
)

particles_sf <- system.file("extdata", "lizard_del_14_1512_sim1_10.json", package = "coralseed") |>
  st_read(quiet = TRUE)
```

## Run a baseline seed-particle model

```{r}

seeded <- seed_particles(
  input = particles_sf,
  zarr = FALSE,
  set.centre = TRUE,
  seascape = seascape,
  probability = "additive",
  limit.time = 12,
  competency.function = "exponential",
  crs = 20353,
  simulate.mortality = "typeIII",
  simulate.mortality.n = 0.1,
  return.plot = TRUE,
  return.summary = TRUE,
  silent = FALSE
)
```

## Inspect outputs

```{r}
names(seeded)

seeded$summary
```

If the returned object includes points, inspect key fields:

```{r}
seeded$points |>
  st_drop_geometry() |>
  glimpse()
```

## Compare competency functions

Use a small scenario grid to compare model behaviour.

```{r}
#| eval: false
competency_models <- c("exponential", "weibull", "lognormal")

seeded_competency <- purrr::map(
  competency_models,
  \(model) {
    seed_particles(
      input = particles_sf,
      zarr = FALSE,
      set.centre = TRUE,
      seascape = seascape,
      probability = "additive",
      limit.time = 12,
      competency.function = model,
      crs = 20353,
      simulate.mortality = "typeIII",
      simulate.mortality.n = 0.1,
      return.plot = FALSE,
      return.summary = TRUE,
      silent = TRUE
    )
  }
) |>
  set_names(competency_models)
```

## Compare mortality curves

```{r}
#| eval: false
mortality_models <- c("typeI", "typeII", "typeIII")

seeded_mortality <- purrr::map(
  mortality_models,
  \(mortality) {
    seed_particles(
      input = particles_sf,
      zarr = FALSE,
      set.centre = TRUE,
      seascape = seascape,
      probability = "additive",
      limit.time = 12,
      competency.function = "exponential",
      crs = 20353,
      simulate.mortality = mortality,
      simulate.mortality.n = 0.1,
      return.plot = FALSE,
      return.summary = TRUE,
      silent = TRUE
    )
  }
) |>
  set_names(mortality_models)
```

## Extract particle tracks

Use `particles_to_tracks()` or related helpers when you need line features for plotting or spatial summaries.

```{r}
#| eval: false
tracks <- particles_to_tracks(seeded$points)
tracks
```
