---
title: "Diagnostics, performance, and pkgdown rendering"
author: "George Roff"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This tutorial gives practical checks for debugging `coralseed` workflows

```{r}
#| eval: false

library(coralseed)
library(tidyverse)
library(sf)
library(ggplot2)
```

## Check spatial inputs

```{r}
#| eval: false

check_sf <- function(x, name = deparse(substitute(x))) {
  cat("\n---", name, "---\n")
  cat("class:", paste(class(x), collapse = ", "), "\n")

  if (!inherits(x, "sf")) {
    cat("not sf\n")
    return(invisible(FALSE))
  }

  cat("nrow:", nrow(x), "\n")
  cat("crs:", sf::st_crs(x)$input, "\n")
  cat("geometry type:", paste(unique(as.character(sf::st_geometry_type(x))), collapse = ", "), "\n")
  cat("empty geometries:", sum(sf::st_is_empty(x)), "\n")
  cat("valid geometries:", sum(sf::st_is_valid(x, NA_on_exception = TRUE), na.rm = TRUE), "\n")

  invisible(TRUE)
}
```

Use it before mapping or modelling:

```{r}
#| eval: false

check_sf(seascape, "seascape")
check_sf(lizard_particles_sf, "particles")
check_sf(land_poly, "land_poly")
```

## Avoid empty-layer map failures

Interactive mapping can fail if an empty `sf` layer reaches `tmap` or `leaflet`. Use small guards inside examples.

```{r}
#| eval: false

valid_sf <- function(x) {
  inherits(x, "sf") &&
    nrow(x) > 0 &&
    length(sf::st_geometry(x)) > 0 &&
    !all(sf::st_is_empty(x))
}
```

```{r}
#| eval: false

if (valid_sf(seascape) && valid_sf(settled$points)) {
  plot_particles(settled$points, seascape)
}
```

## Test with small particle numbers

```{r}
#| eval: false

particle_sim_test <- simulate_particles(
  flow_field = flow_sim$flow_field,
  land_poly = land_poly,
  release_sites = release_sites,
  dt = 5 * 60,
  release_duration_seconds = 20 * 60,
  n_particles_per_site_per_release = 10,
  K = 0.05,
  max_land_retry = 50,
  seed = 101
)
```

## Scale up with parallel release-site simulations

```{r}
#| eval: false

particle_sim <- simulate_particles(
  flow_field = flow_sim$flow_field,
  land_poly = land_poly,
  release_sites = release_sites,
  dt = 5 * 60,
  release_duration_seconds = 60 * 60,
  n_particles_per_site_per_release = 100,
  K = 0.05,
  max_land_retry = 200,
  seed = 101,
  parallel = TRUE,
  workers = 6
)
```

## Simplify land polygons

Detailed coastlines make `st_intersects()` expensive. Simplify land polygons for particle rejection tests.

```{r}
#| eval: false

land_poly_fast <- land_poly |>
  st_transform(32755) |>
  st_make_valid() |>
  st_simplify(dTolerance = 1) |>
  st_transform(4326)
```



## Build locally

```{r}
#| eval: false

devtools::document()
devtools::install(".", upgrade = "never", quick = TRUE)

pkgdown::clean_site(pkg = ".")
pkgdown::build_site(
  pkg = ".",
  new_process = FALSE,
  lazy = FALSE,
  quiet = FALSE
)
```

## Remove render artefacts

```{r}
#| eval: false

unlink("docs", recursive = TRUE, force = TRUE)
unlink("_freeze", recursive = TRUE, force = TRUE)
unlink(".quarto", recursive = TRUE, force = TRUE)
unlink("vignettes/*_files", recursive = TRUE, force = TRUE)
unlink("vignettes/*_cache", recursive = TRUE, force = TRUE)
file.remove(list.files("vignettes", pattern = "\\.md$", full.names = TRUE))
```

